---
title: "Is Democracy in Danger? The Politics of Democratic Backsliding"
subtitle: "Hauptseminar Spring 2026" 
author: 
- 27.02.2026 9:00-15:30 --- INE 220

- 13.03.2026 9:00-17:30 --- INE 220

- 17.04.2026 9:00-17:30 --- INE 220

- 15.05.2026 9:00-15:30 --- 4B02

thanks: "This version of the syllabus: `r format(Sys.time(), '%d %b %Y')`" 

csl: ../00_bibliography/apa.csl
link-citations: true
linkcolor: blue
fontsize: 12pt
number-sections: true
format: 
  pdf:
    include-in-header: 
      text: |
          % LATEX PACKAGES
          \usepackage[T1]{fontenc} % Use the T1 font encoding
          \usepackage{fontspec} % Allow to modify font
          \usepackage{setspace}\onehalfspacing    % Line spacing control
          \usepackage[utf8]{inputenc}  % UTF-8 input encoding
          \usepackage{dcolumn}     % Align on decimal points in  tables
          \usepackage{booktabs}   % Professional tables
          \usepackage{longtable}     % Tables spanning multiple pages
          \usepackage{array}     % Enhanced array and table tools
          \usepackage{multirow}  % Cells spanning multiple rows in  tables
          \usepackage{wrapfig}   % Wrap text around figures
          \usepackage{float}     % Improved float control
          \usepackage{colortbl}  % Color for tables
          \usepackage{pdflscape} % Landscape-oriented pages
          \usepackage{tabu}    % Flexible table environment
          \usepackage{threeparttable}  % Tables with notes
          \usepackage{threeparttablex} % "threeparttable" with  "longtable" support
          \usepackage[normalem]{ulem}  % Various underlining styles
          \usepackage{makecell}         % Enhanced table cell creation
          \usepackage{xcolor}  % Advanced color support
          \usepackage{fancyhdr}    % Custom headers and footers
          \usepackage{enumitem} % itemize
          \usepackage{graphicx} % resize table
          
          % LATEX SETUP
          \pagestyle{fancy}  % Use the fancy header/footer style
          \fancypagestyle{plain}{\pagestyle{fancy}} % Fancy style for plain pages
          \fancyhead[LE,RO]{Dr. Álvaro Canalejo-Molero}
          \fancyhead[RE,LO]{University of Lucerne}

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

# Course description

Over the past decades, growing debates and concerns have emerged about whether we are witnessing a period of democratic decline. These discussions have intensified in recent years, with renewed geopolitical tensions and events such as Donald Trump’s re-election and the rise of the European far right reigniting worries about the stability and resilience of democratic systems. In this context, this course explores the politics of democratic backsliding. It begins by reviewing key theories of democracy and democratic backsliding—understood as processes of state-led democratic erosion—before turning to the potential mechanisms behind these developments. The course first examines backsliding from the perspective of political elites, and then from that of citizens. Finally, it assesses recent trends and debates and concludes with a discussion on the current state of democracy. The aim of the course is to equip students with the conceptual tools to assess democracy and its potential erosion in today’s context, and to engage critically with the latest empirical research at the intersection of democratic theory and political behavior.


# Course organization

This course is organized as a block seminar that will run during the Spring term of 2026. Each block will consist of three or four sessions held on Fridays. The first block provides an overview of theories of democracy, democratic backsliding, and erosion. The second and third blocks then examine the mechanisms sustaining backsliding in today’s societies, focusing on the roles of political elites and citizens in Blocks II and III, respectively. The fourth and final block assesses current trends and debates on the measurement of, and resistance to, democratic backsliding. Students will give at least one presentation in one of the sessions in Blocks II, III, or IV. There will be no student presentations during Block I.

# Enrolment requirements

This course is open to advanced bachelor’s and master’s students interested in democracy and political behavior. Ideally, students should have some basic familiarity with core topics in political science. Since many of the assigned papers include statistical analyses and experimental designs, a basic understanding of statistics (descriptive statistics and regression) is **strongly recommended**.

# Learning outcomes

By the end of this course, students will be able to:

1. **[Knowledge]** Understand key theories of democracy and democratic erosion, and assess major threats, trends, and forms of resilience.

2. **[Competence — Critical reading]** Critically evaluate quantitative research on democracy and political behavior, assessing theory and empirical strategy.

3. **[Competence — Writing]** Write critical response papers on scientific articles related to democracy and political behavior.

4. **[Competence — Communication]** Debate with their colleagues and communicate complex concepts effectively to a broad audience.

# Teaching policy

This course is designed as a seminar in which students are expected to complete the mandatory readings in advance and participate actively during class. Each session consists of two parts. In the first part, the lecturer will introduce the topic based on the readings. In the second part, students will present a case study of a country experiencing democratic backsliding (Blocks II, III, and IV) or take part in a group activity designed to encourage critical thinking (Block I). **Active participation is essential to the success of the course.**

# Integration and interaction policy

Discrimination due to race, gender, ethnicity, or sexual orientation is strictly forbidden in this course. Students are encouraged to use inclusive language and to respect the sexual and gender identification of others. During each session, the lecturer will actively promote equal participation within the class. More generally, interventions must be carried out respectfully, integrating and engaging with different views and perspectives. Discouraging language or other bullying strategies are strictly forbidden, too.

# Artificial intelligence (AI) policy

The use of generative AI tools for preparing class materials is allowed. However, these tools should augment, not replace, human effort (e.g., students should not copy and paste content directly from these tools). The primary concern is not grades but the quality of class discussions, which are central to evaluation. Relying on AI to substitute for human work will hinder meaningful discussion and, ultimately, not only the grade but the learning outcomes. 

To help enforcing this policy, students are required to declare the use of AI and specify its scope in an appendix when submitting response papers, presentation slides or any other assignments.

# Evaluation

## Mandatory requirements (3 credits) {.unnumbered}

To receive the credits, students are expected to fulfill the following criteria:

### 1. Attend all the sessions {.unnumbered}

Attendance is mandatory. Students can miss a maximum of two sessions. Missing more sessions without a justified certificate of absence implies failing the course.

### 2. Do the mandatory readings before each session {.unnumbered}

This course is reading-intensive. Therefore, I recommend *reading cleverly*. For that:

  1. Read the the following blog entry: [How to Read a Scientific Article](https://macartan.github.io/teaching/how-to-read) 

  2. Follow its advice!
  
  3. Do not hesitate to reach out the lecturer with any doubts or questions.

The reading texts will be uploaded to the materials folder in OLAT by the beginning of the course.

### 3. Participate actively {.unnumbered}

Active engagement involves intervening during the discussion with questions, comments and ideas, reflecting on the content of the readings. Students are always encouraged to raise new arguments and debates.

### 4. Write two response papers {.unnumbered}

Students must select two sessions from blocks II, III and IV. For each selected session, they must submit one short response paper (500–700 words).The selected sessions must come from different blocks (e.g., one from Block II and one from Block III), and not belong to the same block as the student’s presentation session.

The response papers should focus on one of the mandatory readings, however, students will benefit from linking them with ideas from additional readings.

A good response paper should:

  - *Start with a short summary of the paper*. It should not be a summary of the content but of the main argument and findings.
  
  - *Identify the contribution.* It should state clearly what bigger problem or question the author is contributing to addressing.
  
  - *Evaluate the argument from a critical perspective.* It should identify the main limitations of the paper and propose ways in which these limitations should be addressed either by the author or in future research.
  
Some strategies for critical reading can be found in this blog entry: [How to Critique a Scientific Article](https://macartan.github.io/teaching/how-to-critique). Follow its advice, and do not hesitate to reach out the lecturer with any doubts or questions.

The **submission rules** are:

- Students must upload their response papers to the *Response papers* folder in OLAT. 

- The deadline for submitting the response paper is at least one week before the corresponding session, that is, the **previous Friday at noon (12:00)**. 

- The file must be in Word. The title should include the session's number followed by the student's surname in capital letters and the response paper number (either 1 or 2). For example: *session05_CANALEJO_1*.

### 5. Present a case study of a country under democratic backsliding  {.unnumbered}

Students will present a case study of a country experiencing democratic backsliding. Presentations will be conducted individually or in pairs and will last **a maximum of 30 minutes**. Case study countries will be **assigned during Block I**.

The objective of the presentation is to analyze democratic backsliding in a concrete national context, combining empirical evidence, institutional analysis, and critical discussion.

#### Structure of the presentation {.unnumbered}

Presentations should be organized around the following four components:

1. **Country background** ***(max. 10 minutes)***

Introduce the country by providing essential contextual information, including:

\begin{itemize}[leftmargin=1.6cm]
  \item Basic socio-economic and geopolitical context
  \item Democratic history (since when the country is considered democratic and to what extent)
  \item Type of democratic system (e.g. presidential or parliamentary; proportional or majoritarian)
  \item Structure of the party system
\end{itemize}

2. **Democratic backsliding and trends over time** ***(max. 10 minutes)***

Analyze democratic developments using empirical evidence, including:

\begin{itemize}[leftmargin=1.6cm]
  \item Data sources measuring democracy over time
  \item When and why the country is considered to be undergoing democratic backsliding
  \item Comparison with neighboring countries
  \item Main dimensions affected (e.g. elections, rule of law, media freedom, checks and balances)
  \item Key milestones, critical events, and the current situation
\end{itemize}

3. **Opposition, resilience, and future prospects** ***(max. 10 minutes)***

Discuss political dynamics and possible trajectories, including:

\begin{itemize}[leftmargin=1.6cm]
  \item Whether and how democratic backsliding is politicized
  \item Organization and strategies of opposition forces
  \item Forms of resistance and institutional or societal resilience
  \item Plausible scenarios for democratic recovery, continued erosion, or democratic breakdown
\end{itemize}

4. **Discussion question**

Conclude with one clear question to open a class discussion. This may involve:

\begin{itemize}[leftmargin=1.6cm]
  \item Evaluating specific developments in the case
  \item Comparing the country with others studied in the course
  \item Discussing alternative strategies or responses
  \item Normative or analytical reflections
\end{itemize}

Creativity is encouraged, as long as the question is clearly connected to the case.

#### Case study countries {.unnumbered}

Case study countries have been selected in advance and assigned to specific sessions.

| Session | Date        | Time          | Case study     |
|--------:|-------------|---------------|----------------|
| 4       | 13.03.2026  | 10:00–10:45   | Hungary        |
| 5       | 13.03.2026  | 12:00–12:45   | Turkey         |
| 6       | 13.03.2026  | 14:45–15:30   | India          |
| 7       | 13.03.2026  | 16:45–17:30   | United States  |
|         |             |               |                |
| 8       | 17.04.2026  | 10:00–10:45   | Bangladesh     |
| 9       | 17.04.2026  | 12:00–12:45   | Nicaragua      |
| 10      | 17.04.2026  | 14:45–15:30   | El Salvador    |
| 11      | 17.04.2026  | 16:45–17:30   | Philippines    |
|         |             |               |                |
| 12      | 15.05.2026  | 10:00–10:45   | Brazil         |
| 13      | 15.05.2026  | 12:00–12:45   | Poland         |


#### Organizational details {.unnumbered}

- Presentation groups and session assignments will be finalized during the introductory session in **Block I**.

- Presentation slides must be uploaded to the *Students responses/Presentation slides* folder in OLAT no later than one day before the session, i.e. **by Thursday at 12:00 (noon)**.

- Slides may be submitted in PowerPoint or PDF format.

- File names must follow this structure: *presentation_slides_sessionXX_SURNAME(S)* (e.g. *presentation_slides_session02_CANALEJO*).

### Grading policy {.unnumbered}

- 10% of the grade is determined by the quality and quantity of the participation in class.

- 40% of the grade is determined by the response papers.

- 50% of the grade is determined by the presentation.

The grade will not only take differences between students into account but also within students' differences over time (i.e., their personal progress).

```{r echo=FALSE, fig.width=6, fig.height=3, fig.align = "center"}
# Load necessary library
library(ggplot2)

# Data for grading policy
grades <- data.frame(
  Category = c("Class Participation", "Response papers", "Presentation"),
  Percentage = c(10, 40, 50)
)

# Reorder the levels of Category based on Percentage from higher to lower
grades$Category <- factor(grades$Category, levels = grades$Category[order(-grades$Percentage)])

# Create a bar chart
ggplot(grades, aes(x = reorder(Category, -Percentage), y = Percentage, fill = Category)) +
  geom_bar(stat = "identity") +
  geom_text(aes(label = paste0(Percentage, "%")), vjust = -0.5) +  # Add percentage labels
  theme_minimal() +
  labs(title = "Grading Policy Breakdown", x = "Category", y = "Percentage") +  # Remove x-axis label
  scale_fill_brewer(palette = "Set2") +
  ylim(0,60) +
  theme(legend.position = "right",  # Add legend for categories
        legend.title = element_blank(),  # Remove legend title
        axis.text.x = element_blank(),  # Remove x-axis text
        plot.title = element_text(hjust = 0.5))  # Center title
```


## Seminar paper (4/6 credits) {.unnumbered}

Students can choose to write a *haupt* (master) seminar paper to obtain four (six) extra credits. The seminar paper must have between 6000 and 10000 words. The topic should be agreed upon between the instructor and the student. To this end, students are asked to write a paper outline of 1-2 pages consisting of the following elements:

-	Introduction of the topic

-	Research question

- Academic and societal relevance

-	Theory and tentative hypotheses

-	Approach and structure of the paper (including a tentative empirical design)

The deadline for the submission of the paper outline is **May 1st 2026**.

An approximate deadline for the submission of the seminar paper is **September 1st 2026**.

Please refer to the [Guidelines on How to Do Research of the Department of Political Science](chrome-extension://efaidnbmnnnibpcajpcglclefindmkaj/https://www.unilu.ch/fileadmin/fakultaeten/ksf/institute/polsem/Dok/Studium/Guidelines_POLSEM___2023___English_01.pdf) for more details.

# Office hours

The lecturer has not fixed office hours. Instead, students can send an e-mail at [alvaro.canalejo@unilu.ch](alvaro.canalejo@unilu.ch) to schedule a meeting within a one-week time period, either in person (office 3.B14) or virtually via Zoom.

# Course schedule

## Block I. Democracy and Democratic Backsliding (27.02.2026 / 9:00–15:30) {.unnumbered}

### Session 1. Introduction (27.02.2026 / 9:00–10:45) {.unnumbered}

**Mandatory reading**

Przeworski, A. (2019). *Crises of democracy*. Cambridge University Press. *Chapter 1*

### Session 2. What is Democracy? (27.02.2026 / 11:00–12:45) {.unnumbered}

**Mandatory readings**

Dahl, R. A. (1972). *Polyarchy: Participation and opposition*. Yale university press. *Chapter 1* 

Przeworski, A. (1991). *Democracy and the market: Political and economic reforms in Eastern Europe and Latin America* (Vol. 181). Cambridge University Press. *Chapter 1*

**Additional readings**

Schmitter, Philippe C., and Terry Lynn Karl. (1991). What Democracy Is…and Is Not. *Journal of Democracy*, 2 (3): 75-88.

Manin, B., Przeworski, A., & Stokes, S. (1999). *Democracy, accountability and representation*. Cambridge University Press. *Chapter 1*

### Session 3. What is Democratic Backsliding? (27.02.2026 / 13:45–15:30) {.unnumbered}

**Mandatory readings**

Bermeo, Nancy. (2016). On Democratic Backsliding. *Journal of Democracy*, 27(1): pp. 5-19.

Waldner, David, and Ellen Lust. (2018). Unwelcome change: Coming to terms with democratic backsliding. *Annual Review of Political Science*, 21: 93-113.

**Additional readings**

Lührmann, A., & Lindberg, S. I. (2019). A third wave of autocratization is here: what is new about it?. *Democratization*, 26(7), 1095-1113.

Carothers, T., & Hartnett, B. (2024). Misunderstanding democratic backsliding. *Journal of Democracy*, 35(3), 24-37.

## Block II. The Supply-Side of Democratic Backsliding (13.03.2026 / 9:00–17:30) {.unnumbered}

### Session 4. Weakening Horizontal Accountability (13.03.2026 / 9:00–10:45) {.unnumbered}

**Mandatory readings**

Varol, Ozan. (2015). Stealth Authoritarianism. *Iowa Law Review*, 100(4): 1673-1742. *Parts I, II and III* 

Levitsky, Steven and Daniel Ziblatt. (2018). *How Democracies Die*. New York: Crown. *Chapter 4* 

**Additional readings**

O’Donnell, Guillermo A. (1994). Delegative Democracy. *Journal of Democracy*, 5 (1): 55-69.

Haggard, S., & Kaufman, R. (2021). *Backsliding: Democratic regress in the contemporary world*. Cambridge University Press. *Chapters 3 & 4* 

### Session 5. Weakening Vertical Accountability (13.03.2026 / 11:00–12:45) {.unnumbered}

**Mandatory readings**

Bentele, Keith G., and Erin E. O’Brien. (2013). Jim Crow 2.0? Why States Consider and Adopt Restrictive Voter Access Policies. *Perspectives on Politics*, 11 (4): 1088-1116. 

Birch, S., & Van Ham, C. (2017). Getting away with foul play? The importance of formal and informal oversight institutions for electoral integrity. *European Journal of Political Research*, 56(3), 487-511.

**Additional reading**

Grossman, Guy, Yotam Margalit, and Tamar Mitts. (2022). How the ultrarich use media ownership as a political investment. *The Journal of Politics*, 84(4): 1913-1931.

Weyland, K. (2025). Opposition to populist backsliding: conditions, limitations, and opportunities. *Democratization*, 32(1), 1-26.

### Session 6. Weakening Democratic Norms (13.03.2026 / 13:45–15:30) {.unnumbered}

**Mandatory readings**

Bursztyn, L., Egorov, G., & Fiorin, S. (2020). From extreme to mainstream: The erosion of social norms. *American Economic Review*, 110(11), 3522-3548.

Clayton, K., Davis, N. T., Nyhan, B., Porter, E., Ryan, T. J., & Wood, T. J. (2021). Elite rhetoric can undermine democratic norms. *Proceedings of the National Academy of Sciences*, 118(23), e2024125118.

**Additional readings**

Valentim, V. (2024). *The normalization of the radical right: A norms theory of political supply and demand*. Oxford University Press. *Chapter 1*

Valentim, V., Dinas, E., & Ziblatt, D. (2025). How mainstream politicians erode norms. *British Journal of Political Science*, 55, e105.

### Session 7. Populism and the Weakening of 'Party Democracy'  (13.03.2026 / 15:45–17:30) {.unnumbered}

**Mandatory readings**

Kriesi, H. (2017). The populist challenge. In *The role of parties in twenty-first century politics* (pp. 131-148). Routledge.

Bessen, B. R. (2024). Populist discourse and public support for executive aggrandizement in Latin America. *Comparative Political Studies*, 57(13), 2118-2151.

**Additional readings**

Mudde, Cas, and Cristobal Rovira Kaltwasser. (2017). *Populism: A Very Short Introduction*. New York: Oxford University Press. *Chapter 1*

Kendall-Taylor, A. and Erica F. (2016). [How Democracies Fall Apart: Why Populism is a Pathway to Autocracy](https://www.foreignaffairs.com/articles/2016-12-05/how-democracies-fall-apart). *Foreign Affairs*
  
## Block III. The Demand-Side of Democratic Backsliding (17.04.2026 / 9:00–17:30) {.unnumbered}

### Session 8. Support for Democracy and Political Trust (17.04.2026 / 9:00–10:45) {.unnumbered}

**Mandatory readings**

Claassen, C. (2020). Does public support help democracy survive?. *American Journal of Political Science*, 64(1), 118-134.

Jacob, M. S. (2025). Citizen support for democracy, anti‐pluralist parties in power and democratic backsliding. *European Journal of Political Research*, 64(1), 348-373.

**Additional readings**

Valgarðsson, V., Jennings, W., Stoker, G., Bunting, H., Devine, D., McKay, L., & Klassen, A. (2025). A crisis of political trust? Global trends in institutional trust from 1958 to 2019. *British Journal of Political Science*, 55, e15.

Claassen, C., Ackermann, K., Bertsou, E., Borba, L., Carlin, R. E., Cavari, A., ... & Zechmeister, E. J. (2025). Conceptualizing and measuring support for democracy: A new approach. *Comparative Political Studies*, 58(6), 1171-1198.

### Session 9. Democratic Values and Hypocrisy (17.04.2026 / 11:00–12:45) {.unnumbered}

**Mandatory readings**

Graham, M. H., & Svolik, M. W. (2020). Democracy in America? Partisanship, polarization, and the robustness of support for democracy in the United States. *American Political Science Review*, 114(2), 392-409.

Chu, J. A., Williamson, S., & Yeung, E. S. (2025). Are people willing to trade away democracy for desirable outcomes? Experimental evidence from six countries. *Comparative Political Studies*, 00104140251392539.

**Additional readings**

Wunsch, N., Jacob, M. S., & Derksen, L. (2025). The demand side of democratic backsliding: How divergent understandings of democracy shape political choice. *British Journal of Political Science*, 55, e39.

Yeung, E. S. F. (2025). Dynamic democratic backsliding. *British Journal of Political Science*, 55, Article e164

### Session 10. Grievances and Resentment (17.04.2026 / 13:45–15:30) {.unnumbered}

**Mandatory readings**

Cramer, K. J. (2016). *The politics of resentment: Rural consciousness in Wisconsin and the rise of Scott Walker*. University of Chicago Press. *Chapter 1*

Berman, Sheri. (2021). The Causes of Populism in the West. *Annual Review of Political Science*, 24 (1), 71-88.

**Additional readings**

Scheiring, G., Serrano-Alarcón, M., Moise, A., McNamara, C., & Stuckler, D. (2024). The populist backlash against globalization: A meta-analysis of the causal evidence. *British Journal of Political Science*, 54(3), 892-916.

Margalit, Y., Raviv, S., & Solodoch, O. (2025). The cultural origins of populism. *The Journal of Politics*, 87(2), 393-410.

### Session 11. Media Change and Disinformation (17.04.2026 / 15:45–17:30) {.unnumbered}

**Mandatory readings**

Lecheler, S., & Egelhofer, J. L. (2022). Disinformation, misinformation, and fake news: Understanding the supply side. In *Knowledge resistance in high-choice information environments* (pp. 69-87). Taylor & Francis

Tucker, J. A., Guess, A., Barberá, P., Vaccari, C., Siegel, A., Sanovich, S., ... & Nyhan, B. (2018). Social media, political polarization, and political disinformation: A review of the scientific literature. *Working paper* (March 19, 2018) 

**Additional readings**

Jerit, J., & Zhao, Y. (2020). Political misinformation. *Annual Review of Political Science*, 23(1), 77-94.

Ecker, U., Roozenbeek, J., Van Der Linden, S., Tay, L. Q., Cook, J., Oreskes, N., & Lewandowsky, S. (2024). Misinformation poses a bigger threat to democracy than you might think. *Nature*, 630(8015), 29-32.

## Block IV. Trends and Debates on Democratic Backsliding (15.05.2026 / 15:45–17:30) {.unnumbered}

### Session 12. Measuring Democratic Backsliding (15.05.2026 / 9:00–10:45) {.unnumbered}

**Mandatory readings**

Marina, N., Lundstedt, M., Altman, D., Angiolillo, F., Borella, C., Fernandes, T., Gastaldi, L., Good, A. G., Natsika N., & Lindberg, S. I. (2024). Democracy Report 2024: Democracy Winning and Losing at the Ballot. *University of Gothenburg: V-Dem Institute* 

Little, A. T., & Meng, A. (2024). Measuring democratic backsliding. *PS: Political Science & Politics*, 57(2), 149-161.

**Additional readings**

Knutsen, C. H., Marquardt, K. L., Seim, B., Coppedge, M., Edgell, A. B., Medzihorsky, J., ... & Lindberg, S. I. (2024). Conceptual and measurement issues in assessing democratic backsliding. *PS: Political Science & Politics*, 57(2), 162-177.

Baron, H., Blair, R. A., Gottlieb, J., & Paler, L. (2024). An events-based approach to understanding democratic erosion. *PS: Political Science & Politics*, 57(2), 208-215.

### Session 13. Institutional and Civil Resistance (15.05.2026 / 11:00–12:45) {.unnumbered}

**Mandatory readings**

Graham, B. A., Miller, M. K., & Strøm, K. W. (2017). Safeguarding democracy: Powersharing and democratic survival. *American Political Science Review*, 111(4), 686-704.

Chenoweth, E., & Stephan, M. J. (2011). *Why civil resistance works: The strategic logic of nonviolent conflict*. Columbia University Press. *Chapter 1* 

**Additional readings**

Capoccia, G. (2026). Countering Illiberalism in Liberal Democracies: Information, Legacies, Temporalities. *Comparative Political Studies*, 0(0). 00104140261418663

Bateman, D. A. (2025). Democracy-reinforcing hardball: Can breaking democratic norms preserve democratic values?. *Comparative Political Studies*. 00104140241312107.

### Session 14. Is Democracy in Danger? Final Remarks (15.05.2026 / 13:45–15:30) {.unnumbered}

**Mandatory reading**

Riedl, R. B., Friesen, P., McCoy, J., & Roberts, K. (2024). Democratic backsliding, resilience, and resistance. *World Politics*.

**Additional readings**
 
Samuels, D. J. (2023). The international context of democratic backsliding: Rethinking the role of third wave “prodemocracy” global actors. *Perspectives on Politics*, 21(3), 1001-1012.  
 
Benmelech, E., & Monteiro, J. (2026). War and Democratic Backsliding (No. w34734). *National Bureau of Economic Research*. 
 
Pattison, J. (2025). The ethics of responding to democratic backsliding abroad. *American Journal of Political Science*.

